---
title: "orderly"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{orderly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r echo = FALSE, results = "hide"}
lang_output <- function(x, lang) {
  writeLines(c(sprintf("```%s", lang), x, "```"))
}
r_output <- function(x) lang_output(x, "r")
yaml_output <- function(x) lang_output(x, "yaml")
plain_output <- function(x) lang_output(x, "plain")
orderly_file <- function(...) {
  system.file(..., package = "orderly", mustWork = TRUE)
}

path <- orderly::orderly_init(tempfile(), quiet = TRUE)
path_example <- file.path(path, "src", "example")
dir.create(path_example)
orderly:::file_copy(orderly_file("orderly_config_sqlite.yml"),
                    file.path(path, "orderly_config.yml"), overwrite = TRUE)
orderly:::file_copy(orderly_file(c("report_example.yml", "report_example.R")),
                    file.path(path_example, c("orderly.yml", "script.R")))
orderly:::orderly_default_config_set(orderly:::orderly_config(path))
local({
  con <- orderly::orderly_db("source")
  DBI::dbWriteTable(con, "mtcars", mtcars)
})
orderly::orderly_log_start()

tree <- function(path, header = path) {
  paste1 <- function(a, b) {
    paste(rep_len(a, length(b)), b)
  }
  indent <- function(x, files) {
    paste0(if (files) "| " else "  ", x)
  }
  is_directory <- function(x) {
    unname(file.info(x)[, "isdir"])
  }
  sort_files <- function(x) {
    i <- grepl("^[A-Z]", x)
    c(x[i], x[!i])
  }
  prefix_file <- "|--="
  prefix_dir  <- "|-+="

  files <- sort_files(dir(path))
  files_full <- file.path(path, files)
  isdir <- is_directory(files_full)

  ret <- as.list(c(paste1(prefix_dir, files[isdir]),
                   paste1(prefix_file, files[!isdir])))
  files_full <- c(files_full[isdir], files_full[!isdir])
  isdir <- c(isdir[isdir], isdir[!isdir])

  n <- length(ret)
  if (n > 0) {
    ret[[n]] <- sub("|", "\\", ret[[n]], fixed = TRUE)
    tmp <- lapply(which(isdir), function(i)
      c(ret[[i]], indent(tree(files_full[[i]], NULL), !all(isdir))))
    ret[isdir] <- tmp
  }

  c(header, unlist(ret))
}
```

## The problem

Suppose that we have an SQL database that is used for a series of
"reports" (e.g., graphs, knitr documents, data exports).  These
reports might change because:

* the data in the SQL database changes
* we apply the reporting scripts to different subsets of the data
* the code that we use to generate the scripts changes

(among other more pathalogical reasons such as the packages that
we're using changing behaviour, etc)

We would like to be able to generate reports easily from the
database and later on compare two reports and easily be able to
determine _why_ two versions of a report differ - particularly with
respect to the above criteria.  We'd also like to be able to get
the original data that was used to create the reports (even if the
SQL database has moved on) and the original scripts that were used
to create the reports.

## Example

The `orderly.yml` to describe the creation of a report might look like:
``` {r results = "asis", echo = FALSE}
yaml_output(readLines(file.path(path_example, "orderly.yml")))
```

Hopefully this is somewhat self explanatory:

* `data` is the data sets to pull from the database; it is a named
  list with one or more elements.  Each element is a SQL query
  indicating the required data.  You can use placeholders like
  `?cyl` in the above example to stand in for *parameters* that
  will be passed through to the report.

* `parameters` is a list of parameters used to query the SQL
  database.  Note the that only the *names* of the parameters are
  provided here - the values will be provided when the report is
  run.

* `script.R` is the R script that will be run to do the actual work
  of the report - any R code can be used here.

* `artefacts` is a list of "artefacts" that are created by the
  `script`; these are *files* that contain the report itself.
  There can be more than one artefact (there are two here).  Under
  each filename, list the `format` (`"staticgraph"`,
  `"interactivegraph"`, `"data"` for now) and a free-text
  description.

There are other optional fields that can be added to the yaml (and
`parameters` is optional) but this is the core.  The file
`script.R` for this report contains

``` {r results = "asis", echo = FALSE}
yaml_output(readLines(file.path(path_example, "script.R")))
```

The important thing here is that the report references the variable
`cars` and `cyl` even though it does not create them!  It must
produce the files that are mentioed in `artefacts`
(`disp_vs_wt.png` and `distribution.png`).  The R code can be as
long and as short as needed and can use whatever packages it needs.
`orderly` does not do anything with the script apart from run it so
it can be formatted however (there are no magic comments, etc).
There is no restrictions on what can be done except:

* don't try to directly access the database
* the script must create the artefacts listed in `orderly.yml`

There is some more infrastructure that needs to be put in place
around this; how does the report access the SQL database?  And
where to the generated reports reside?
``` {r results = "asis", echo = FALSE}
plain_output(tree(path, "<root>"))
```

The directories will be discussed in further detail below but we have:

* `archive` - completed reports.  These are reports that we might
  distribute to people or synchronise with a central point.
* `data` - copies of the data used to create reports, stored so
  that reports can be rerun later.
* `draft` - draft reports.  Every run of a report will create a new
  directory in here, but relatively few of these will be "final"
  reports.  So reports exist here before being copied to `archive`.
* `src` - the report *sources*; the files `orderly.yml` and
  `script.R` are the files from above.
* `orderly_config.yml` - this is the global configuration
* `source.sqlite` - this is the example database used here

The global configuration looks like:
``` {r results = "asis", echo = FALSE}
yaml_output(readLines(file.path(path, "orderly_config.yml")))
```

This defines the field `source`, which describes the database that
we pull results from (the database that the report queries are run
against).  This contains a field `driver` which must declare a
DBI-compatible driver (realistically this is going to be
`RSQLite::SQLite` for a local SQLite database and
`RPostgres::Posgres` for accessing a postgres database), and then
any arguments to be passed through to `DBI::dbConnect` along with
the driver.  For SQLite this is just going to be `dbname` but for
postgres this will include `host`, `port`, `user`, `dbname` and
`password`.  For example:

```yaml
source:
  driver: RSQLite::SQLite
  host: dbhost
  port: 5432
  user: myusername
  password: s3cret
  dbname: mydb
```

In order to run anything (as below) the working directory must be
somewhere above the root.  `orderly` will look in directories up
towards the filesystem root for a file `orderly_config.yml` to
determine the root.  For Rstudio users, creating an .Rproj file
here is probably a good idea.

## Running a report

Reports are run by *name*.  In the above configuration the report
is called `example` because it is in the directory `src/example`.
A list of report names can be returned by running
``` {r }
orderly::orderly_list()
```

If the report requires parameters then these must be passed through
too; in the above case the report takes the parameter `cyl` which
must be passed through as a named list.  So we might run this
report as
``` {r collapse = TRUE}
id <- orderly::orderly_run("example", list(cyl = 4))
```

``` {r echo = FALSE, results = "hide"}
h <- sub("\\.rds$", "", dir(file.path(path, "data", "rds")))
```

The return value is the id of the report (also printed on the
second line of log output) and is always in the format
`YYYYMMDD-HHMMSS-abcdef01` where the last 8 characters are a hex
digits (i.e., 4 random bytes).  This means reports will
automatically sort nicely but we'll have some collision resistance.
``` {r }
id
```

Having run the report, the directory layout now looks like:
``` {r results = "asis", echo = FALSE}
plain_output(tree(path, "<root>"))
```

Within `drafts`, the directory ``r file.path("example", id)`` has been
created which contains the result of running the report.  In here
there are the files:

* `orderly.yml`: this is an exact copy of the input file
* `script.R`: this is an exact copy of the script used for the analysis
* `disp_vs_wt.png` and `distribution.png`: the artefacts created by
   the report
* `orderly_run.yml`: this is metadata about the run (hash of input
  files, of the data used, and of the output etc)
* `orderly_run.rds`: version information on all R packages used (in
  R's rds format as I don't think we need this from anything but R)

Every time a report is run it will create a new directory at this
level with a new id.  Running the report again now might create the
directory ``r file.path("example", orderly:::new_report_id())``

Note that there are other files created;

* ``r file.path("data", "csv", paste0(h, ".csv"))``
* ``r file.path("data", "rds", paste0(h, ".rds"))``

These are a copy of the data as extracted from the database and
used in the report.  The filenames are derived from a hash of the
contents of the file, so if two reports use the same data then only
one copy will be saved (this is quite likely if the upstream data
and the query have not changed).  Two copies are saved - the rds
version is faster to read and write, usually smaller, and is
exactly the data as used, while the csv version is more easily read
by other applications.

You can see the list of draft reports like so:
``` {r }
orderly::orderly_list_drafts()
```

Once you're happy with a report, then "commit" it with
``` {r collapse = TRUE}
orderly::orderly_commit(id)
```

**THIS WILL CHANGE A LITTLE I THINK** - but mostly in how the index
is built and how we might synchronise reports across people and
machines.

After this step our directory structure looks like:
``` {r results = "asis", echo = FALSE}
plain_output(tree(path, "<root>"))
```

Which looks very like the previous, but files have been moved from
being within `draft` to being within `archive`.  The other
difference is that the index `orderly.sqlite` has been created.

## Developing a report

First, create a directory within `src`.  The name is important and
should not contain spaces (nor should it change as this will change
the key report id and you'll lose a chain of history).
``` {r }
dir.create(file.path(path, "src", "new"))
```
